### 2023-12-23 ms.qmd
Update ms.qmd

Cognition can be defined as the mechanisms by which animals acquire, process, and store information from their environment through perception, learning, memory, and decision making (@shettleworth_2009); and it underpins several aspects of animals' ecology such as habitat selection, foraging decisions, mate choice, antipredatory strategies, and/or social behaviours (@dukas_evolutionary_2004). Among cognitive traits, learning represents the acquisition of neuronal representations of new information (@dukas_evolutionary_2004) and is, as such, essential for dealing with environmental changes (@dukas_evolutionary_2004; @leal_behavioural_2012; @buchanan_condition_2013).

Effects of maternal effects and early environment on learning. Why temperature and GCs?

Cognition it is affected by the conditions in which organisms develop (Burger, 1991; Lemaire et al., 2000; Amiel and Shine, 2012; Amiel et al., 2014; Abayarathna and Webb, 2020).

Among these effects, maternal glucocorticoids (GCs) have been demonstrated to influence offspring phenotype across taxa (Lemaire et al., 2000; Belliure et al., 2004; Haussmann et al., 2012; Monaghan, 2014; Owen et al., 2018; Haq et al., 2021). For instance, restraint stress in pregnant rats leads to impairments in neurogenesis and learning (Lemaire et al., 2000; Zhu et al., 2004); and direct experimental elevation of maternal GCs can affect morphology and behaviour of lizards (Belliure et al., 2004; Owen et al., 2018).
In this vein, the temperature at which embryos develop (also referred as early thermal environment) is a pervasive mechanism of phenotypic variation in ectotherms (Noble et al., 2017); having substantial and unpredictable effects on diverse traits such as growth (Shine et al., 1995), behaviour (Burger, 1989; 1990; 1991; Dayananda and Webb, 2017; Vila-Pouca et al., 2018; 2019), and survival (Burger, 1989; Vila-Pouca et al., 2019).

How are temperature and GCs related to each other?
Metabolic effects, effects on the brain morphology and function.

Objectives, hypothesis and predictions

Here, we investigated how prenatal Corticosterone (CORT) - the main GCs in reptiles - and incubation temperature interact to shape learning abilities in two species of skinks, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). We experimentally increased CORT levels in the eggs of these two species of skinks and then incubated them at two different temperatures in a 2X2 factorial design. Then, we subjected the juveniles to a colour-associative and a reversal task to explore their learning skills. We hypothesized that changes in CORT levels and temperature during development will have sustained effects on brain's morphology, physiology, and gene expression, ultimately affecting learning skills. We predicted that individuals exposed to high levels of CORT and/or low temperatures will perform worse in the learning tasks than control individuals or individuals exposed to high temperatures. We also expected incubation at high temperatures to decrease the impact of CORT on learning, while cold incubation temperatures were expected to increase CORT detrimental effects.

@@

The subjects in our experiments were juveniles from two species of skinks, the common garden skink (*Lampropholis guichenoti*) and the rainbow skink or delicate skink (*L. delicata*). These are small (∼35--55 mm snout-vent length (SVL)), oviparous, and generalist skinks that usually share the same habitat in suburban areas throughout south-eastern Australia ([@chapple_know_2011]). They also have similar breeding periods, but with some differences in reproductive output: while *L. delicata* lays 1 to 6 eggs in only one clutch per season, *L. guichenoti* clutches are smaller (1-5 eggs per clutch) but they usually lay two clutches per season ([@chapple_know_2011]; [@chapple_biology_2014]). This difference in reproductive output through the breeding season may involve different degree of sensitivity to maternal effects and its interaction with other aspects of early environment.

@@

*Breeding colony* -- We tested juveniles coming from a breeding colony established in the lab since 2019. There is a total of 270 and 180 adults of *L. delicata* and *L.* guichenoti respectively, housed in big containers (41.5 L x 30.5 W x 21 H cm) with six lizards (2 males and 4 females) per enclosure. Enclosures are provided with non-stick matting, shelter, and several small water dishes. Water is given daily, and they are fed approx. 40 mid-size crickets (*Acheta domestica*) per enclosure three days a week. Crickets are dusted with calcium weekly and multivitamin and calcium biweekly. To ensure a temperature gradient, we employ a heat chord and a heat lamp following a 12 h light:12 h dark cycle. Room temperatures are set to 22-24 Celsius, and warm side of enclosures is usually at 32 Celsius.

>>>>

Cognition is defined as the processes by which animals gather, preserve, and use information from their environment through perception, learning, memory, and decision making (@shettleworth_2009). These mechanisms underpin several aspects of animals' ecology such as habitat selection, foraging decisions, mate choice, antipredatory strategies, and/or social behaviours that are crucial for survival and reproduction (@dukas_evolutionary_2004). Particularly, learning - the acquisition of neuronal representations of new information (@dukas_evolutionary_2004) - is seen as fundamental for coping with environmental changes by allowing individuals to create new associations between events (@dukas_evolutionary_2004; @leal_behavioural_2012; @buchanan_condition_2013). However, the ability of organisms to acquire new information varies accross individuals, with this natural variation usually being influenced by factors like age, sex, social system, gut microbiota, or the environment where animals develop (**REFS**). In this regard, the latter can be particularly relevant because the brain is especially sensitive to environmental impacts during early stages of development (**REFS**). 
These differences in neural development caused by prenatal environment can have long-lasting effects on learning skills (**REFS**). In this regard, prenatal Glucocorticoids (GCs) and prenatal thermal environment are known to play a prominent role in shaping learning abilities in different taxa. GCs - **hormones related to organisms' response to changes in homeostasis (BEST DEF AND REF**) - can have sustained pysiological effects on neural structure and physiology associated with animals' performance on learning tasks (**Definition and REF**). In this sense, there is evidence that high prenatal GC levels impair with learning accross different taxa (**MORE EG.S AND REFS**; @lemaire_prenatal_2000; @zhu_prenatal_2004), but results vary depending on sex, or task (**REFS**). Also, some studies have revealed effects of prenatal temperature on learning in ectotherms (**REFS**). For instance, ... (**REFS**). In this vein, the effect of prenatal temperature has been associated with similar changes in neural structure and metabolic activity (**REFS**). However, despite both prenatal GCs and thermal environment are known to affect learning through similar mechanisms, the interaction between this two factors is still poorly understood.
 In this study, we aim to explore the interactive effects between prenatal Glucocorticoids (GCs) and the prenatal thermal environment on learning using two species of skinks, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_), as our model species. We experimentally increased Corticosterone (CORT) - the main GC in reptiles - levels in the eggs of these two species of skinks and then incubated them at two different temperatures in a 2X2 factorial design. Then, we subjected the juveniles to a colour-associative and a reversal task to explore their learning abilities. We hypothesise that changes in CORT levels and temperature during early development will have sustained effects on brain's morphology and physiology that will ultimately impact learning skills. We predicted that individuals exposed to high levels of CORT and/or low temperatures will perform worse in the learning tasks than control individuals or individuals exposed to high temperatures. We also expected incubation at high temperatures to decrease the impact of CORT on skink's performance, while cold incubation temperatures were expected to increase CORT detrimental effects on learning.

**Cognition it is affected by the conditions in which organisms develop (Burger, 1991; Lemaire et al., 2000; Amiel and Shine, 2012; Amiel et al., 2014; Abayarathna and Webb, 2020).**

@@

_L. guichenoti_ and _L. delicata_ are small (∼35--55 mm snout-vent length (SVL)), oviparous, and generalist skinks that usually share the same habitat in suburban areas throughout south-eastern Australia ([@chapple_know_2011]). They also have similar breeding periods, but with some differences in reproductive output: while *L. delicata* lays 1 to 6 eggs in only one clutch per season, *L. guichenoti* clutches are smaller (1-5 eggs per clutch) but they usually lay two clutches per season ([@chapple_know_2011]; [@chapple_biology_2014]).

@@

*Breeding colony* -- We tested juveniles coming from a breeding colony established in the lab since 2019. There is a total of 270 and 180 adults of *L. delicata* and *L. guichenoti* respectively, housed in big containers (41.5 L x 30.5 W x 21 H cm) with six lizards (2 males and 4 females) per enclosure. Enclosures are provided with non-stick matting, shelter, and several small water dishes. Water is given daily, and they are fed approx. 40 mid-size crickets (*Acheta domestica*) per enclosure three days a week. Crickets are dusted with calcium weekly and multivitamin and calcium biweekly. To ensure a temperature gradient, we employ a heat chord and a heat lamp following a 12 h light:12 h dark cycle. Room temperatures are set to 22-24 Celsius, and warm side of enclosures is usually at 32 Celsius.
